Financialisation of Commodity Markets \\\large{Co-movement Behind-the-Scenes}

### 2025-08-25 docs/paper.tex
Update body of text
for bar charts and EW portfolios table.

Table~\ref{tbl-stats-combined} displays descriptive statistics for the
commodity series considered in the study; the results show a clear
pattern of increase in average return during financialisation with an
equally weighted portfolio of the US traded assets showing a mean
returns of 15.9\% significant at the 5\% level vs.~6.9\% with a 10\%
significance in the pre-financialisation period. The increase is even
more pronounced for an equally weighted portfolio of GB traded assets
(metals), 2.47\% (not significant) to 21.8\% (significant at the 10\%
level), which sets out the global nature of the phenomenon. Most
commodities show a higher mean return while five---Lumber (XCME),
Natural gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary
(XLME)---show a lower figure with five og them significant---Crude
oil-WTI (XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and
Tin-refined (XLME)---against none pre-financialisation. Sugar-\#11 wins
the race with a mean return increasing from -4.6\% to 24.9\% followed by
copper with sizeable increases for both markets: 1.5\% to 28.2\% (XCEC)
and 0.5\% to 28.6\% (XLME).\\
The pattern of results for volatilities is similar although the increase
is less pronounced. Both equally weighted portfolios shows an increase,
more pronounced for the GB assets, 10\% to 13.6\% (US) vs.~15.7\% to
26.9\% (GB), while out of the thirty assets, eleven show a drop in
volatility: Oats (XCBT), Lean hogs (XCME), Cocoa (IFUS), Coffee-C
(IFUS), Lumber (XCME), Sugar-\#11 (IFUS), Natural gas (XNYM), Crude
oil-WTI (XNYM), Gasoline (XNYM), Heating oil (XNYM), Palladium (XNYM).
Financialisation seems however to have impacted the commodity complex
beyond these mechanical effects of increased long participation ensuing
from the entry on index type investments as our results suggest next.\\
The cross-sectional nature of financialisation naturally leads us to
extend the individual hedging pressure paradigm by incorporating
aggregate CHP (\textbf{CHP}), a market wide measure of hedging pressure.
The traditional Keynesian hedging pressure paradigm postulates a
negative relationship between the return on an individual commodity
futures and its hedging pressure. The extended Keynesian hedging
pressure paradigm, in the context of financialisation, should also imply
that returns for individual commodities should be high in periods where
\textbf{CHP} is low and vice et versa. Table~\ref{tbl-stats-combined}
further shows the mean returns and volatilities for high
(\textbf{contango}) and low (\textbf{backwardation}) \textbf{CHP}
regimes for each time period while
Table~\ref{tbl-regime-diference-tests} shows the results of difference
tests for returns mean and variance between the two regimes.\\
Over the first period eighteen of the twenty-four US commodities have
higher mean returns during periods of \textbf{backwardation} as compared
to \textbf{contango}, with the difference being statistically
significant at the 10\% level for two commodities, at the 5\% level for
one and at the 1\% level for another. For the remaining six commodities
the higher mean return during \textbf{contango} is not statistically
significant for any of them. For the six GB commodities, all metals,
five of them had higher mean returns during phases of
\textbf{backwardation} as compared to \textbf{contango} in the first
period, with one of these differences being statistically significant at
the 10\% level while for the remaining asset (Zinc (XLME)) the higher
mean return during \textbf{contango} is not statistically significant.
An equally weighted portfolio of the twenty-four US commodities had a
mean return of 15.4\% during \textbf{backwardation} phases,
statistically significant at the 1\% level, and a mean return of -1.1\%
during \textbf{contango}, both in the first period, with the difference
between them statistically significant at the 5\% level. The
corresponding figures for an equally weighted portfolio of the GB metals
were 4.4\% and 0.3\% respectively. These results provide broad support
for an aggregate version of the Keynesian hedging pressure hypothesis
and suggest that hedgers in the aggregate were engaged in risk transfer
during this period.\\
Over the second period the pattern essentially reverses. For the
twenty-four US commodities over this period, fourteen have higher mean
returns during phases of \textbf{contango} over \textbf{backwardation}.
Four of the six GB metals achieved higher mean returns over
\textbf{contango} relative to \textbf{backwardation} phases with the
difference being statistically significant at the 5\% level for one
commodity (Nickel-primary (XLME)). The equally weighted US portfolio had
a mean return of 15.7\% over \textbf{backwardation} against 16.2\%
during phases of \textbf{contango}, both statistically significant at
the 10\% level. The difference was considerably more pronounced for the
equally weighted GB metals portfolio with a mean return of 8\% during
\textbf{backwardation} rising to 35.6\% during \textbf{contango}.

@@

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities, eventually spurring the transformation of commodity markets
from a collection of idiosyncratic markets into a more cohesive
collection akin to an asset class. In this context, interconnectedness
emerges as a crucial aspect that warrants thorough examination.\\
We strive to address this issue by first examining the average returns
pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector.
Table~\ref{tbl-correlations-inner-periods} shows the results for the
four periods of interest with the results reported for the entire period
as well as for periods of \textbf{backwardation} and \textbf{contango}.
For the entire sample the average pairwise correlation jumps from 0.07
in the first period to 0.16 over financialisation. This finding
indicates higher interdependence within the full cross-section of
commodities starting with the financialisation period, consistent with
the empirical findings in \citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods. In this context,
the ontological consequences of financialisation are better observed
from an investment perspective as illustrated in
Table~\ref{tbl-correlations-cross-periods} where we observe average
pairwise correlation coefficients between equally weighted portfolios of
US traded commodity assets grouped by sub-sector\footnote{A more
  detailed breakdown of country, sector and subsector average pairwise
  correlations is provided in the attached
  \href{http://bautheac.duckdns.org:3838/bautheac/revision-jfm/}{Internet
  Appendix}.}. Although modest in magnitude, these exhibit a telling
pattern, mirroring that shown by the individual commodity returns in
Table~\ref{tbl-stats-combined}, with a substantially higher average
coefficient over \textbf{backwardation} phases in the first period (0.12
vs.~0.03) and the pattern reversed over financilisation with an average
coefficient substantially higher in \textbf{contango} (0.12 vs.~0.02).\\
We break down Table~\ref{tbl-correlations-inner-periods}'s average
correlations by year in Table~\ref{tbl-correlations-inner-years}. For
the full sample of assets, the average pairwise correlations remain in
the 0.06--0.08 range from 1998 through 2003, before suddenly increasing
to 0.12 from 2003 to 2004. This suggests 2004 as a distinct break point
for the onset of financialisation, as noted in several studies
\citep{baker_financialization_2021, tang_index_2012}. We further observe
that the US metals appear to have played a major role in this as the
increase seems driven by the sector which average pairwise correlation
more than doubles (0.49 vs 0.22) over the period.

@@

\textbf{backwardation}. The onset of financialisation leads to a
substantial improvement in the performance of the CHP factor with its
average \(R^{2}\), albeit still small at 5.4\%, almost tripling from the
past to the financialisation period in contrast to the other factors
that show weaker dynamics with a slight decrease and a much smaller
increase for the open interest and term structure factors respectively.
Here again the increase in pricing performance for the CHP factor is
driven by the metals sector, particularly the precious sub-sector with
the average \(R^{2}\) for the group jumping to almost 26\% during
financialisation from 1.2\% over the past period. The corresponding
figures of 22\% and 1.1\% for the entire US metals sector stand in sharp
contrast with the agricultural sector, which exhibits no notable change,
and the energy sector, where pricing performance drops substantially
from 9.2\% down to 1.6\%. The improvement in the CHP factor's pricing
performance remarkably comes predominantly from the long leg (low CHP
assets: Keynesian ``normal backwardation'') of the factor\footnote{Results
  available from the authors upon request.} with the corresponding
mimicking portfolio's performance itself driven by the metals, the
precious sub-sector in particular. Interestingly, although the term
structure factor itself only sees a very modest increase, its short leg
(low roll-yield assets: Working's backwardation) shows a sizeable rise
in pricing performance, yet again driven the metals sector. The pattern
of results for both factors extends to the GB traded assets and suggest
a convergence of Keneysian and Working's term structure perspective over
the financialisation period.\\

@@

The crisis period (late 2008 to mid-2013) shows a contrasting pattern of
results. Returns fall sharply relative to the financialisation period
(Table~\ref{tbl-stats-combined}) with eighteen US commodities along with
all six GB traded metals showing lower mean returns. Returns for equally
weighted portfolios of both sets of assets fall down to 9.4\% and 5.3\%
from 15.9\% and 21.8\% respectively. Returns are mostly higher during
phases of \textbf{backwardation} relative to \textbf{contango} with
fifteen US assets and five of the GB traded metals showing higher mean
returns in \textbf{backwardation} and the pattern extending to equally
weighted portfolios of the two sets of assets. Volatility is moderately
higher for both sets with fifteen US commodities and three GB metals
showing higher figures and interestingly eminently higher over phases of
\textbf{contango}; out of the US complex only Sugar-\#11 (IFUS) and none
of the GB traded metals show higher volatility figures over
\textbf{backwardation} phases. The symmetry between the dynamics of CHP
and \textbf{CHP}'s explanatory power on individual commodity returns
breaks (Table~\ref{tbl-time-reg-US-CHP}) with both now seemingly
converging; the US cross-sectional average \(R^{2}\) at the individual
level bouncing back up (22.5\% vs 20.8\%) while that at the aggregate
level rises further (14.7\% vs 7.1\%).

@@

The returns pull-back continues into the post-crisis period
(Table~\ref{tbl-stats-combined}) with twenty-two US and four GB traded
assets showing lower mean returns relative to the crisis while the same
number of US commodities and all GB metals show lower figures relative
to the financialisation period; mean returns for equally weighted
portfolios of US and GB assets drop down to -0.2\% and 2.7\%
respectively. Interestingly volatility also recedes substantially with
twenty US assets and all GB metals showing lower volatility figures
relative to both the crisis and financialisation periods suggesting that
investment outflows from the commodity futures complex have had a lower
impact on volatility than the corresponding inflows witnessed over the
financialisation period. The figures for the equally weighted portfolios
of US and GB assets drop to 9.9\% and 16.2\% respectively, very close to
pre-financialisation levels.\\
The pattern of reversion toward the pre-financialisation Keynesian
paradigm initiated during the crisis continues stronger into the
post-crisis period with mean returns higher during phases of
\textbf{backwardation} relative to \textbf{contango}, more so than over
the crisis period with seventeen US commodities showing higher mean
returns over \textbf{backwardation} phases. The difference
(Table~\ref{tbl-regime-diference-tests}) is significant at the 5\% level
for Lumber (XCME) and at the 10\% level for Palladium (XNYM) while no
difference is significant for the assets showing higher figures over
phases of \textbf{contango}. The pattern is more pronounced for the GB
traded subset with all six assets showing higher mean returns in
\textbf{backwardation} with the difference significant at the 1\% level
for Nickel-primary (XLME), 5\% level for Aluminium (XLME) and 10\% level
for Zinc (XLME). It also extends to equally weighted portfolios of US
and GB traded assets that both show higher mean returns during
\textbf{backwardation} with the difference significant at the 5\% level
for the GB portfolio and the figure close to the 10\% level significance
threshold for the portfolio of US assets. Volatilities are overall
higher in \textbf{contango} although less so than over the previous
period with seventeen US commodities showing higher figures over
\textbf{contango} against twenty-three in the crisis period. The pattern
extends to three of the six GB traded metals with the difference
significant at the 1\% level for Copper (XLME) and Tin (XLME) and at the
5\% level for Zinc (XLME) while differences are not significant for
Aluminium (XLME), Lead (XLME) and Nickel-primary (XLME) that exhibit
higher volatility over \textbf{backwardation} phases. Equally weighted
portfolios of US and GB traded assets on the other hand both show higher
volatility in \textbf{contango} with the difference significant at the
1\% level for both. Over the financialisation period in contrast
volatilities were predominantly higher in \textbf{backwardation} for US
traded assets with fourteen commodities showing higher volatility
figures in these phases; the post-crisis pattern of results for
volatilities thus likewise shows signs of reversion to
pre-financialisation standards. After showing signs of convergence in
the crisis period the dynamics of the explanatory power of CHP and
\textbf{CHP} part ways (Table~\ref{tbl-time-reg-US-CHP}), seemingly
reverting to the Keynesian paradigm observed pre-financialisation,
further suggesting a return to legacy commodity specific dynamics. The
recovery initiated in the crisis period for that of CHP is confirmed
with a stronger increase, 27.2\% vs.~22.5\%, strongest for the metals
sector, 37.3\% vs.~23.7\%, while the energy sector shows the opposite
pattern: 11\% vs.~16\%. In contrast, after increasing stronger in the
crisis, that for \textbf{CHP} falls overall post-crisis, 7.5\% vs
14.7\%, with the decrease strongest for the energy sector in a relative
sense: 2.5\% vs.~14.8\%.

@@

\newgeometry{margin=.75in}

@@

\caption{\label{tbl-stats-combined}This table shows mean returns and
corresponding volatility for the twenty four individual US commodities
and the six GB metals considered in the study as well as for two equally
weighted portfolios formed from the US commodities and the GB metals
respectively across the four periods of interest (past: 1997-2003;
financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018).
for each period the figures are shown independently for the entire
period as well as for phases of aggregate backwardation (aggregate CHP
\(\leq\) period median) and aggregate contango (aggregate CHP
\textgreater{} period median). Mean values significant at the 1\%, 5\%
and 10\% level are marked with ***, ** and * respectively. Aggregate CHP
construction and corresponding regime definitions are discussed in
Section~\ref{sec-methods} while the results are discussed in
Section~\ref{sec-results}.}

@@

\caption{\label{tbl-time-reg-US-CHP}This table shows the average time
series \(R^{2}\) for models where the returns series for the twenty four
individual US commodities are independently regressed against relative
change in their own commercial hedging pressure (CHP) series as well as
against that in aggregate CHP, a market wide measure of CHP which
calculation method is discussed in Section~\ref{sec-methods}. Averages
are calculated independently across the whole cross-section of US traded
commodity assets considered in the study as well as across commodity
sectors and sub-sectors for the four periods of interest (past:
1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis:
2013-2018). The results are discussed in Section~\ref{sec-results} while
Section~\ref{sec-taxonomy} details the taxonomy for the commodity assets
considered.}

@@

\caption{\label{tbl-correlations-inner-periods}This table shows the
average return pairwise correlation coefficients calculated
independently across the whole cross-section of commodity assets
considered in the study as well as within countries and commodity
sectors and sub-sectors for the four periods of interest (past:
1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis:
2013-2018). For each period, the results are calculated independently
over the entire period as well as over phases of aggregate backwardation
and contango. Aggregate CHP construction and corresponding regime
definitions are discussed in Section~\ref{sec-methods} while the results
are discussed in Section~\ref{sec-results} and the taxonomy for the
commodity assets considered is detailed in Section~\ref{sec-taxonomy}.}

@@

\caption{\label{tbl-correlations-cross-periods}This table shows the
average return pairwise correlation coefficients between equally
weighted US sub-sector portfolios (agriculturals: gains, livestock,
softs; energy: gas, petroleum; metals: base, precious). The results are
independently calculated for the four periods of interest (past:
1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis:
2013-2018). For each period, the results are further calculated
independently over the entire period as well as over phases of aggregate
backwardation and contango. Aggregate CHP construction and corresponding
regime definitions are discussed in Section~\ref{sec-methods} while the
results are discussed in Section~\ref{sec-results} and the taxonomy for
the commodity assets considered is detailed in
Section~\ref{sec-taxonomy}.}

@@

\toprule
\textcolor{black}{\textbf{regime}} & \textcolor{black}{\textbf{past}} & \textcolor{black}{\textbf{financialisation}} & \textcolor{black}{\textbf{crisis}} & \textcolor{black}{\textbf{post-crisis}}\\
\midrule
\textbf{whole period} & \textcolor[HTML]{4285f4}{0.0421} & \textcolor[HTML]{4285f4}{0.0631} & \textcolor[HTML]{4285f4}{0.0799} & \textcolor[HTML]{4285f4}{0.0778}\\
\textbf{backwardation} & \textcolor[HTML]{4285f4}{0.1166} & \textcolor[HTML]{4285f4}{0.0201} & \textcolor[HTML]{4285f4}{0.0633} & \textcolor[HTML]{4285f4}{0.0727}\\
\textbf{contango} & \textcolor[HTML]{4285f4}{0.0297} & \textcolor[HTML]{4285f4}{0.1164} & \textcolor[HTML]{4285f4}{0.0889} & \textcolor[HTML]{4285f4}{0.0812}\\
\bottomrule

@@

\begin{longtable}[t]{>{}l>{}l>{}l>{}l>{}l>{}r>{}r>{}r>{}r}

>>>>

The results show a clear pattern of increase in average return during
financialisation as illustrated by Figure~\ref{fig-estimates}, with most
individual commodities showing a higher mean return, and
Table~\ref{tbl-stats-regime-tests} with an equally weighted portfolio of
the US traded assets showing a mean returns of 15.9\% significant at the
5\% level vs.~6.9\% with a 10\% significance in the pre-financialisation
period. The increase is even more pronounced for an equally weighted
portfolio of GB traded assets (metals), 2.47\% (not significant) to
21.8\% (significant at the 10\% level), which sets out the global nature
of the phenomenon. The pattern of results for volatilities is similar
although the increase is less pronounced with a weaker majority of
individual commodities showing heightened volatility and both equally
weighted portfolios showing an increase, more pronounced for the GB
assets. Financialisation seems however to have impacted the commodity
complex beyond these mechanical effects of increased long participation
ensuing from the entry of index type investments as our results suggest
next.\\
The cross-sectional nature of financialisation naturally leads us to
extend the individual hedging pressure paradigm by incorporating
aggregate CHP (\textbf{CHP}), a market wide measure of hedging pressure.
The traditional Keynesian hedging pressure paradigm postulates a
negative relationship between the return on an individual commodity
futures and its hedging pressure. The extended Keynesian hedging
pressure paradigm, in the context of financialisation, should also imply
that returns for individual commodities should be high in periods where
\textbf{CHP} is low and vice et versa.\\
Over the first period twenty three of the thirty commodity assets
considered have a higher mean returns during periods of
\textbf{backwardation} as compared to \textbf{contango}
(Figure~\ref{fig-regimes}) including eighteen US and five GB traded
assets. For the US commodities the difference is statistically
significant at the 10\% level for two assets, at the 5\% level for one
and at the 1\% level for another; for the remaining six commodities the
higher mean return during \textbf{contango} is not statistically
significant for any of them. For the GB commodities, the difference is
significant at the 10\% level for one asset while for the remaining
asset the higher mean return during \textbf{contango} is not
statistically significant. An equally weighted portfolio of the
twenty-four US commodities had a mean return of 15.4\% during
\textbf{backwardation} phases, statistically significant at the 1\%
level, and a mean return of -1.1\% during \textbf{contango}, both in the
first period, with the difference between them statistically significant
at the 5\% level. The corresponding figures for an equally weighted
portfolio of the GB metals were 4.4\% and 0.3\% respectively. These
results provide broad support for an aggregate version of the Keynesian
hedging pressure hypothesis and suggest that hedgers in the aggregate
were engaged in risk transfer during this period.\\
Over the second period the pattern essentially reverses with seventeen
of the thirty commodities having higher mean returns during phases of
\textbf{contango} over \textbf{backwardation} including fourteen US and
four GB traded assets with the difference being statistically
significant at the 5\% level for one commodity. The equally weighted US
portfolio had a mean return of 15.7\% over \textbf{backwardation}
against 16.2\% during phases of \textbf{contango}, both statistically
significant at the 10\% level. The difference was considerably more
pronounced for the equally weighted GB metals portfolio with a mean
return of 8\% during \textbf{backwardation} rising to 35.6\% during
\textbf{contango}.

@@

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities, eventually spurring the transformation of commodity markets
from a collection of idiosyncratic markets into a more cohesive
collection akin to an asset class. In this context, interconnectedness
emerges as a crucial aspect that warrants thorough examination.\\
We strive to address this issue by first examining the average returns
pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector.
Table~\ref{tbl-correlations-inner-periods} shows the results for the
four periods of interest with the results reported for the entire period
as well as for periods of \textbf{backwardation} and \textbf{contango}.
For the entire sample the average pairwise correlation jumps from 0.07
in the first period to 0.16 over financialisation. This finding
indicates higher interdependence within the full cross-section of
commodities starting with the financialisation period, consistent with
the empirical findings in \citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods. In this context,
the ontological consequences of financialisation are better observed
from an investment perspective as illustrated in
Table~\ref{tbl-correlations-cross-periods} where we observe average
pairwise correlation coefficients between equally weighted portfolios of
US traded commodity assets grouped by sub-sector\footnote{A more
  detailed breakdown of country, sector and subsector average pairwise
  correlations is provided in the attached
  \href{https://bautheac.shinyapps.io/co-movement/}{Internet Appendix}.}.
Although modest in magnitude, these exhibit a telling pattern, mirroring
that shown by the individual commodity returns in
Table~\ref{tbl-stats-regime-tests}, with a substantially higher average
coefficient over \textbf{backwardation} phases in the first period (0.12
vs.~0.03) and the pattern reversed over financilisation with an average
coefficient substantially higher in \textbf{contango} (0.12 vs.~0.02).\\
We break down Table~\ref{tbl-correlations-inner-periods}'s average
correlations by year in Table~\ref{tbl-correlations-inner-years}. For
the full sample of assets, the average pairwise correlations remain in
the 0.06--0.08 range from 1998 through 2003, before suddenly increasing
to 0.12 from 2003 to 2004. This suggests 2004 as a distinct break point
for the onset of financialisation, as noted in several studies
\citep{baker_financialization_2021, tang_index_2012}. We further observe
that the US metals appear to have played a major role in this as the
increase seems driven by the sector which average pairwise correlation
more than doubles (0.49 vs 0.22) over the period.

@@

\textbf{backwardation}\footnote{A breakdown of the performance by regime
  is provided in the attached
  \href{https://bautheac.shinyapps.io/co-movement/}{Internet Appendix}.}.
The onset of financialisation leads to a substantial improvement in the
performance of the CHP factor with its average \(R^{2}\), albeit still
small at 5.4\%, almost tripling from the past to the financialisation
period in contrast to the other factors that show weaker dynamics with a
slight decrease and a much smaller increase for the open interest and
term structure factors respectively. Here again the increase in pricing
performance for the CHP factor is driven by the metals sector,
particularly the precious sub-sector with the average \(R^{2}\) for the
group jumping to almost 26\% during financialisation from 1.2\% over the
past period. The corresponding figures of 22\% and 1.1\% for the entire
US metals sector stand in sharp contrast with the agricultural sector,
which exhibits no notable change, and the energy sector, where pricing
performance drops substantially from 9.2\% down to 1.6\%. The
improvement in the CHP factor's pricing performance remarkably comes
predominantly from the long leg (low CHP assets: Keynesian ``normal
backwardation'') of the factor\footnote{A breakdown of the performance
  by factor leg is provided in the attached
  \href{https://bautheac.shinyapps.io/co-movement/}{Internet Appendix}.}
with the corresponding mimicking portfolio's performance itself driven
by the metals, the precious sub-sector in particular. Interestingly,
although the term structure factor itself only sees a very modest
increase, its short leg (low roll-yield assets: Working's backwardation)
shows a sizeable rise in pricing performance, yet again driven the
metals sector. The pattern of results for both factors extends to the GB
traded assets and suggest a convergence of Keneysian and Working's term
structure perspective over the financialisation period.\\

@@

The crisis period (late 2008 to mid-2013) shows a contrasting pattern of
results. Returns fall sharply relative to the financialisation period
(Figure~\ref{fig-estimates}) with twenty-four commodities showing lower
mean returns, including eighteen US and six GB traded assets. Returns
for equally weighted portfolios of both sets of assets
(Table~\ref{tbl-stats-regime-tests}) fall down to 9.4\% and 5.3\% from
15.9\% and 21.8\% respectively. Returns are mostly higher during phases
of \textbf{backwardation} relative to \textbf{contango}
(Figure~\ref{fig-regimes}) with twenty commodities showing higher mean
returns in \textbf{backwardation}, including fifteen US and five of the
GB traded assets, and the pattern extending to equally weighted
portfolios of the two sets of assets. Volatility is moderately higher
with eighteen assets showing higher figures, fifteen US commodities and
three GB metals; and interestingly eminently higher over phases of
\textbf{contango}; out of the US complex only Sugar-\#11 (IFUS) and none
of the GB traded metals show higher volatility figures over
\textbf{backwardation} phases. The symmetry between the dynamics of CHP
and \textbf{CHP}'s explanatory power on individual commodity returns
breaks (Table~\ref{tbl-time-reg-US-CHP}) with both now seemingly
converging; the US cross-sectional average \(R^{2}\) at the individual
level bouncing back up (22.5\% vs 20.8\%) while that at the aggregate
level rises further (14.7\% vs 7.1\%).

@@

The returns pull-back continues into the post-crisis period with twenty
six commodities (Figure~\ref{fig-estimates}), including twenty-two US
and four GB traded assets, showing lower mean returns relative to the
crisis while the same number of US commodities and all GB metals show
lower figures relative to the financialisation period; mean returns for
equally weighted portfolios of US and GB assets
(@Table~\ref{tbl-stats-regime-tests}) drop down to -0.2\% and 2.7\%
respectively. Interestingly volatility also recedes substantially with
twenty six commodities, including twenty US assets and all GB metals,
showing lower volatility figures relative to both the crisis and
financialisation periods suggesting that investment outflows from the
commodity futures complex have had a lower impact on volatility than the
corresponding inflows witnessed over the financialisation period. The
figures for the equally weighted portfolios of US and GB assets drop to
9.9\% and 16.2\% respectively, very close to pre-financialisation
levels.\\
The pattern of reversion toward the pre-financialisation Keynesian
paradigm initiated during the crisis continues stronger into the
post-crisis period with mean returns higher during phases of
\textbf{backwardation} relative to \textbf{contango}
(Figure~\ref{fig-regimes}), more so than over the crisis period with
twenty three commodities showing higher mean returns over
\textbf{backwardation} phases. Seventeen of those are US traded assets
with the difference significant at the 5\% level for Lumber (XCME) and
at the 10\% level for Palladium (XNYM) while no difference is
significant for the assets showing higher figures over phases of
\textbf{contango}. The pattern is more pronounced for the GB traded
subset with all six assets showing higher mean returns in
\textbf{backwardation} with the difference significant at the 1\% level
for Nickel-primary (XLME), 5\% level for Aluminium (XLME) and 10\% level
for Zinc (XLME). It also extends to equally weighted portfolios of US
and GB traded assets that both show higher mean returns during
\textbf{backwardation} with the difference significant at the 5\% level
for the GB portfolio and the figure close to the 10\% level significance
threshold for the portfolio of US assets. Volatilities are overall
higher in \textbf{contango} although less so than over the previous
period with twenty one commodities showing higher figures over
\textbf{contango} against twenty-eight in the crisis period, including
seventeen US traded assets. The pattern extends to three of the six GB
traded metals with the difference significant at the 1\% level for
Copper (XLME) and Tin (XLME) and at the 5\% level for Zinc (XLME) while
differences are not significant for Aluminium (XLME), Lead (XLME) and
Nickel-primary (XLME) that exhibit higher volatility over
\textbf{backwardation} phases. Equally weighted portfolios of US and GB
traded assets on the other hand both show higher volatility in
\textbf{contango} with the difference significant at the 1\% level for
both. Over the financialisation period in contrast volatilities were
predominantly higher in \textbf{backwardation} for US traded assets with
fourteen commodities showing higher volatility figures in these phases;
the post-crisis pattern of results for volatilities thus likewise shows
signs of reversion to pre-financialisation standards. After showing
signs of convergence in the crisis period the dynamics of the
explanatory power of CHP and \textbf{CHP} part ways
(Table~\ref{tbl-time-reg-US-CHP}), seemingly reverting to the Keynesian
paradigm observed pre-financialisation, further suggesting a return to
legacy commodity specific dynamics. The recovery initiated in the crisis
period for that of CHP is confirmed with a stronger increase, 27.2\%
vs.~22.5\%, strongest for the metals sector, 37.3\% vs.~23.7\%, while
the energy sector shows the opposite pattern: 11\% vs.~16\%. In
contrast, after increasing stronger in the crisis, that for \textbf{CHP}
falls overall post-crisis, 7.5\% vs 14.7\%, with the decrease strongest
for the energy sector in a relative sense: 2.5\% vs.~14.8\%.

@@

\section*{Figures}\label{figures}
\addcontentsline{toc}{section}{Figures}

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-estimates-1.pdf}

}

\caption{\label{fig-estimates}This figure compares the evolution of
estimates for mean (top panel) and volatility (bottom panel) for
individual commodity returns across three period transitions: past →
financialisation, financialisation → crisis and crisis → post-crisis.
Each stacked bar shows how many cases fall into two categories: yellow
(the estimate in the previous period is smaller than in the next period)
and blue (the estimate in the previous period is larger than in the next
period); numbers inside the bars indicate the count of observations in
each category. Detailed results are available in the attached
\href{https://bautheac.shinyapps.io/co-movement/}{Internet Appendix}.
Period definitions are discussed in Section~\ref{sec-methods} while the
results are discussed in Section~\ref{sec-results}.}

\end{figure}%

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-regimes-1.pdf}

}

\caption{\label{fig-regimes}This figure compares estimates for mean (top
panel) and volatility (bottom panel) for individual commodity returns
between \textbf{CHP} regime phases (\textbf{backwardation}
vs.~\textbf{contango}) for the four periods. Each stacked bar shows how
many cases fall into two categories: yellow (the estimate for the
\textbf{backwardation phase} is smaller than for the \textbf{contango}
phase) and blue (the estimate for the \textbf{backwardation phase} is
greater than for the \textbf{contango} phase); numbers inside the bars
indicate the count of observations in each category.
\textbf{Backwardation} and \textbf{contango} regimes as well as period
definitions are discussed in Section~\ref{sec-methods} while the results
are discussed in Section~\ref{sec-results}.}

\end{figure}%

@@

\caption{\label{tbl-stats-regime-tests}This table shows mean returns and
volatilities for two equally weighted portfolios of US and GB traded
assets for the four periods. For each period, estimates are computed
independently over the whole period as well as over phases of
\textbf{backwardation} and \textbf{contango}. The table further shows
the results of t-tests of significance for the mean returns as well as
t-tests of difference between these when computed over the two phases
independently along with F-tests of difference for the corresponding
variances. Values significant at the 1\%, 5\% and 10\% level are marked
with ***, ** and * respectively. Detailed results are available in the
attached \href{https://bautheac.shinyapps.io/co-movement/}{Internet
Appendix}. \textbf{Backwardation} and \textbf{contango} regimes as well
as period definitions are discussed in Section~\ref{sec-methods} while
the results are discussed in Section~\ref{sec-results}.}

@@

\caption{\label{tbl-time-reg-US-CHP}This table shows the average time
series \(R^{2}\) for models where the returns series for the individual
US traded commodities are independently regressed against relative
change in their own commercial hedging pressure (CHP) series as well as
against that in \textbf{CHP}, a market wide measure of CHP which
calculation method is discussed in Section~\ref{sec-methods}. Averages
are computed independently across the whole cross-section of US traded
commodity assets as well as across commodity sectors and sub-sectors for
the four periods. Period definitions are discussed in
Section~\ref{sec-methods} while the results are discussed in
Section~\ref{sec-results} and the commodity assets taxonomy is detailed
in Section~\ref{sec-taxonomy}.}

@@

\caption{\label{tbl-correlations-inner-periods}This table shows the
average return pairwise correlation coefficients computed independently
across the whole cross-section of individual commodity assets as well as
within countries and commodity sectors and sub-sectors for the four
periods. For each period, the results are computed independently over
the entire period as well as over phases of \textbf{backwardation} and
\textbf{contango}. \textbf{Backwardation} and \textbf{contango} regimes
as well as period definitions are discussed in Section~\ref{sec-methods}
while the results are discussed in Section~\ref{sec-results} and the
commodity assets taxonomy is detailed in Section~\ref{sec-taxonomy}.}

@@

\caption{\label{tbl-correlations-cross-periods}This table shows the
average return pairwise correlation coefficients between equally
weighted US sub-sector portfolios (agriculturals: gains, livestock,
softs; energy: gas, petroleum; metals: base, precious) independently
computed for the four periods. For each period, the results are further
computed independently over the entire period as well as over phases
\textbf{backwardation} and \textbf{contango}. \textbf{Backwardation} and
\textbf{contango} regimes as well as period definitions are discussed in
Section~\ref{sec-methods} while the results are discussed in
Section~\ref{sec-results} and the commodity assets taxonomy is detailed
in Section~\ref{sec-taxonomy}.}

@@

\toprule
\textcolor{black}{\textbf{regime}} & \textcolor{black}{\textbf{past}} & \textcolor{black}{\textbf{financialisation}} & \textcolor{black}{\textbf{crisis}} & \textcolor{black}{\textbf{post-crisis}}\\
\midrule
\textbf{whole period} & \textcolor[HTML]{4285f4}{\textbf{0.0421}} & \textcolor[HTML]{4285f4}{\textbf{0.0631}} & \textcolor[HTML]{4285f4}{\textbf{0.0799}} & \textcolor[HTML]{4285f4}{\textbf{0.0778}}\\
\textbf{backwardation} & \textcolor[HTML]{4285f4}{\textbf{0.1166}} & \textcolor[HTML]{4285f4}{\textbf{0.0201}} & \textcolor[HTML]{4285f4}{\textbf{0.0633}} & \textcolor[HTML]{4285f4}{\textbf{0.0727}}\\
\textbf{contango} & \textcolor[HTML]{4285f4}{\textbf{0.0297}} & \textcolor[HTML]{4285f4}{\textbf{0.1164}} & \textcolor[HTML]{4285f4}{\textbf{0.0889}} & \textcolor[HTML]{4285f4}{\textbf{0.0812}}\\
\bottomrule

@@

\begin{longtable}[t]{>{}l>{}l>{}l>{}l>{}r>{}r>{}r>{}r}
